Compositional Certified Resource Bounds\ifshort{\vspace{-.15cm}}{}

### 2014-11-11 paper/main.tex
start work on soundness proof

The soundness of the analysis builds on the quantitative logic.  We
translate judgements of the automatic amortized analysis as defined in
\pref{fig:auto} into quantitative Hoare triples.  We define a
translation function $\tr$, such that if a judgement $J$ in the
automatic analysis is derivable, $\tr(J)$ is derivable in the
quantitative logic.  By using $\tr$ to translate derivations of the
automatic analysis to derivations in the quantitative logic we can
automatically obtain a certified resource bound for the analyzed
program.

The translation of an assertion $(\Gamma, Q)$ in the automatic
analysis is defined by
$$
  \tr(\Gamma, Q) := \lambda\state .\, (\Gamma(\state)) + \Phi_Q(\state),
$$
where we write $\Phi_Q$ for the unique linear potential function
defined by the quantitative annotation $Q$.  We also need to
translate the assumptions in function contexts of the automatic
analysis.  We define $\tr(\Gamma_f, Q_f, \Gamma'_f, Q'_f) := \forall z \, \vec v \, v . $
%
$$
(
  \lambda \_ \, \vec v \, \state .\, \tr(\Gamma_f, Q_f)(\state[\Vargs {\mapsto} \vec v]),
  \lambda \_ \, v \, \state .\, \tr(\Gamma'_f, Q'_f)(\state[\Vret {\mapsto} v])
  )
$$
%
These definitions let us translate the judgement $J = B; R; P \vdash
S \dashv P'$ in the context $\Delta$ by
$$
  \tr(J) :=
  (\lambda f .\, \tr(\Delta(f))); \tr(B); \tr(R) \vdash
    \htriple{\tr(P)}{S}{\tr(P')}.
$$
The soundness of the automatic analysis can now be stated
formally with the following theorem.
%
\begin{theorem}[Soundness of the automatic analysis]
  If $J$ is a judgement derived with the rules of
  \pref{fig:auto}, then $\tr(J)$ is a quantitative
  Hoare triple derivable with the rules of
  \pref{fig:logic}.
\end{theorem}
%
\noindent
The proof of this theorem is constructive and basically maps each rule
of the automatic analysis directly to its counterpart in the
quantitative logic.  The most tricky parts are the translations of the
rules for increments and decrements and the rule {\sc Q:Weak} for
weakening because we have to show that the preconditions of the rules
{\sc L:Weak} or {\sc L:Update}, respectively, are met.  \iffull{ For instance
we prove the following lemma to show the soundness of the translation
of {\sc Q:Weak}.

\begin{lemma}[Relax]
  If $Q' \succeq_\Gamma Q$ and $\state$ is a program state such that
  $\state \models \Gamma$ then we have $\Phi_{Q'}(\state) \ge \Phi_Q(\state)$.
\end{lemma}
\begin{proof}
We observe the following inequations.
\begin{align*}
\Phi_{Q'}(\state) &= \sum_i q'_i f_i(\state) \\
&\ge q_0 + \sum_{i\in\mathcal U} u_i p_i - \sum_{i\in\mathcal L} l_i p_i
  + \sum_{i\in\mathcal U} (q_i - p_i) f_i(\state)  \\ &
  \;\;\;\;\;\;\;\; + \sum_{i\in\mathcal L} (q_i + p_i) f_i(\state)
  + \sum_{i\not\in \mathcal U \cup \mathcal L} q_i \\
 &\ge \Phi_Q(\state)
  + \sum_{i\in\mathcal U} (u_i - f_i(\state)) p_i
  + \sum_{i\in\mathcal L} (f_i(\state) - l_i) p_i \\
 &\ge \Phi_Q(\state).
\end{align*}
Since $\state \models \Gamma$, using the transitivity of $\models$ and
the definition of $\mathcal U$ and $\mathcal L$ we get
$\forall i\in\mathcal L.\, l_i \le f_i(\state)$ and
$\forall i\in\mathcal U.\, f_i(\state)\le u_i$.
These two inequalities justify the last step.
\end{proof}
%
\noindent
We prove similar lemmas for the rules that deal with assignment.  With
the computational content of the proof, we have an effective algorithm
to construct certificates for upper bounds derived automatically.
}{}

@@

\paragraph{Semantics.}

CompCert Clight's operational semantics is based on small-step
transitions and continuations.  Expressions---which do not have side
effects---are evaluated in a big-step fashion.  Here, we describe a
simplified version of Clight's semantics for the subset we consider.

A program state $\state = (\theta, \gamma)$ is composed of two maps from
variable names to integers. The first map, $\theta : \text{Locals} \to
\Z$, assigns integers to local variables of a function, and the second
map, $\gamma : \text{Globals} \to \Z$, gives values to global variables
of the program.  In this article, we assume that all values are
integers but in the implementation we support all data types of
Clight.
%
The evaluation function \evalE{\cdot}{} maps an expression $e \in
E$ to a value $\evalE{e}{\state} \in \Z$ in the program state
$\state$.

For simplicity, we assume that local variables and global variables
are always different.  We just write $\state(x)$ to obtain the value of
$x$ in program state $\state$. \iffull{Such a lookup is defined as
$$
(\theta, \gamma)(x) =
\left\{
\begin{array}{lr}
\theta(x) & \mbox{ if } x \in \dom \theta \\
\gamma(x) & \mbox{ if } x \in \dom \gamma
\end{array}
\right..
$$}{}
\iffull{For a program state $\state$, we}{Similarly} we write $\state[x \mapsto v]$
for the program state that maps $x$ to $v$ and behaves as $\state$ for
all other variables, regardless whether $x \in \dom \theta$ or $x \in
\dom \gamma$.

The small-step semantics is standard, except that it tracks the
resource consumption of a program.  The semantics is parametric in the
resource of interest for the user of our system.  We achieve this
independence by parameterizing evaluations with a resource
\emph{metric} $M$; a tuple of nine rational numbers \ifshort{$M_a$, $M_b$, $M_r$, $\ldots$}{}, one map $M_e :
\text{Expr} \to \Q$ from expressions to rational numbers, and one map
$M_t : \Z \to \Q$ from integers to rational numbers.
%
\iffull{$$
M = (
M_a,
M_b,
M_r,
M_u,
M_f,
M_l,
M_c^1,
M_c^2,
M_s,
M_e(E),
M_t(n)
)
$$}{

}
Each of these rational numbers indicates the amount of resource
consumed by a corresponding operation.  If the assigned resource cost
is negative then it means that resources are released.  The metric
that we use in the implementation can also depend on other information
that is statically available such as the name of the called function
or the number of arguments.

\pref{fig:opsem} contains \ifshort{selected}{the} reduction rules of the semantics.  The
rules define a rewrite system for program configurations of the form
$(\state, S, \cont, \cost)$, where $\state$ is the program state, $S$
is the statement being executed, $\cont$ is a continuation that
describes what remains to be done after the execution of $S$, and
$\cost \in \Q$ is the non-negative number of resources available for
further execution.  All rules that can decrease $\cost$ have the
implicit side condition that the resource quantity available
\emph{before} the step is non-negative.  This means that we allow
$(\state, S, \cont, \cost)$ with $c<0$ on the right-hand side
transition relation $\smallstep{M}^*$ to indicate that the execution
ran out of resources.  However, every execution that reaches such a
state is stuck.

\iffull{
A continuation $\cont$ represents the context of the execution of a
given statement.  A continuation can be the empty continuation
$\Kstop$ (used to start a program), a sequenced computation $\Kseq S
\cont$, a loop continuation $\Kloop S \cont$, or the continuation
$\Kcall r \theta \cont$ of a function call.
$$
\cont := \Kstop
\mid \Kseq S \cont
\mid \Kloop S \cont
\mid \Kcall r \theta \cont
$$
%
During the execution we assume a fixed function context $\fenv$ that
maps function names to a list of variables and the statement that
defines the function body.  It is used in the rule \textsc{S:Call}.
}{}

The intuitive meaning of an evaluation \iffull{judgement}{} $(\state,
S, \cont, \cost) \smallstep{M}^* (\state', S', \cont', \cost')$ is the
following.  If the statement $S$ is executed in program state
$\state$, with continuation $\cont$, and with $\cost$ resources
available then---after a finite number of steps---the evaluation will
reach the new machine state $(\state', S', \cont',\cost')$ and there
are $\cost'$ resources available.  If $\cost' \ge 0$ then the
execution did not run out of resources and the resource consumption up
to this point is $\cost - \cost'$.  If this difference is negative
then resource became available during the execution.  If however
$\cost' < 0$ then the execution ran out of resources and is stuck.
The cost of the execution is then $\cost \ge 0$.

\iffull{
\pref{lem:sem1} summarizes the main properties of the resource
counter of the semantics.  It can be proved by induction on the number
of steps.
\begin{lemma}
\label{lem:sem1}
\itemskip
  \begin{enumerate}
  \item If $(\state, S, \cont, \cost) \smallstep{M}^k (\state', S',
    \cont', \cost')$ and $n\ge 0$ then $(\state, S, \cont, \cost {+} n)
    \smallstep{M}^k (\state', S', \cont', \cost' {+} n)$.
\itemskipIn
  \item If $(\state, S, \cont, \cost) \smallstep{M}^k (\state', S',
    \cont', \cost')$ and $(\state, S, \cont, d) \smallstep{M}^k (\state', S',
    \cont', d')$ then $\cost - \cost' = d - d'$.
  \end{enumerate}
\itemskip
\end{lemma}
%
If we assume that we have a \emph{positive resource metric}, that is,
a metric that assigns a positive cost to each evaluation step then we
can link bounds on the resource consumption under the metric to
termination.  This is formalized by \pref{lem:sem_term}.
%
\begin{lemma}
 \label{lem:sem_term}
 Let $M$ be a positive metric and assume that there exists $b \in
 \Qplusz$ such that $(\state, S, \cont, \cost) \smallstep{M}^*
 (\state', S', \cont', \cost') \implies \cost {-} \cost' {\leq} b$.
 Then there exists $B \in \N$ such that $(\state, S, \cont, d)
 \smallstep{M}^n (\state'', S', \cont', d')$ for some $n$, $d$, $d'$
 and $\state''$ implies $n \le B$.
\end{lemma}
}
{
The extended version of this article~\cite{anon_extended} contains all rules and
lemmas that state the main properties of the cost semantics.
}

>>>>

The soundness of the analysis builds on a new cost semantics for Clight
and an extended Quantitative Logic.  Using these two tools, the soundness
proof of \toolname is done by a exhibiting a translation morphism.

\subsection{Cost Semantics for Clight}

To base the soundness proof on a formal ground, we start by defining a
new cost-aware operational semantics for Clight.  CompCert Clight's
operational semantics is based on small-step transitions and
continuations.  Expressions---which do not have side effects---are
evaluated in a big-step fashion.

A program state $\state = (\theta, \gamma)$ is composed of two maps
from variable names to integers. The first map, $\theta : \text{Locals}
\to \Z$, assigns integers to local variables of a function, and the
second map, $\gamma : \text{Globals} \to \Z$, gives values to global
variables of the program.  In this article, we assume that all values are
integers but in the implementation we support all data types of Clight.
The evaluation function \evalE{\cdot}{} maps an expression $e \in E$
to a value $\evalE{e}{\state} \in \Z$ in the program state $\state$.
% For simplicity, we assume that local variables and global variables
% are disjoint.
We write $\state(x)$ to obtain the value of $x$ in program state
$\state$.
% \iffull{Such a lookup is defined as
% $$
% (\theta, \gamma)(x) =
% \left\{
% \begin{array}{lr}
% \theta(x) & \mbox{ if } x \in \dom \theta \\
% \gamma(x) & \mbox{ if } x \in \dom \gamma
% \end{array}
% \right..
% $$}{}
Similarly, we write $\state[x \mapsto v]$ for the state based on $\state$
where the value of $x$ is updated to $v$.

The small-step semantics we define is standard, except that it tracks
the resource consumption of a program.  The semantics is parametric in
the resource of interest for the user of our system.  We achieve this
independence by parameterizing evaluations with a resource \emph{metric}
$M$; a tuple of nine rational numbers and two maps.  Each of these
parameters indicates the amount of resource consumed by a corresponding
step in the semantics.  Resources can be released by assigning a negative
cost. %  The metric used in the implementation can also
% depend on other information that is statically available such as the
% name of the called function or the number of arguments.
Two rules for update and tick follow.
\begin{mathpar}
\Rule[rightskip=0.3cm]{U}
{ \state' = \state[x \mapsto \sem{e}_\state] }
{ (\state, x \gets e, \cont, \cost)
\rightarrow\\\\
(\state', \code{skip}, \cont, \cost {-} M_u {-} M_e(e))
}
\and
\Rule[leftskip=0.3cm,rightskip=0.2cm]{T}
{ }
{ (\state, \code{tick}(n), \cont, \cost)
\rightarrow\\\\
(\state, \code{skip}, \cont, \cost {-} M_t(n))
}
\end{mathpar}
The rules keep track of the resource cost $\cost$ at all times, and
all rules have as implicit side condition that $\cost$ in the start
configuration is non-negative.  It makes it possible
to detect a resource crash as a stuck configuration where $\cost<0$.

% \pref{fig:opsem} contains \ifshort{selected}{the} reduction rules of the semantics.  The
% rules define a rewrite system for program configurations of the form
% $(\state, S, \cont, \cost)$, where $\state$ is the program state, $S$
% is the statement being executed, $\cont$ is a continuation that
% describes what remains to be done after the execution of $S$, and
% $\cost \in \Q$ is the non-negative number of resources available for
% further execution.  All rules that can decrease $\cost$ have the
% implicit side condition that the resource quantity available
% \emph{before} the step is non-negative.  This means that we allow
% $(\state, S, \cont, \cost)$ with $c<0$ on the right-hand side
% transition relation $\smallstep{M}^*$ to indicate that the execution
% ran out of resources.  However, every execution that reaches such a
% state is stuck.
%
% \iffull{
% A continuation $\cont$ represents the context of the execution of a
% given statement.  A continuation can be the empty continuation
% $\Kstop$ (used to start a program), a sequenced computation $\Kseq S
% \cont$, a loop continuation $\Kloop S \cont$, or the continuation
% $\Kcall r \theta \cont$ of a function call.
% $$
% \cont := \Kstop
% \mid \Kseq S \cont
% \mid \Kloop S \cont
% \mid \Kcall r \theta \cont
% $$
% %
% During the execution we assume a fixed function context $\fenv$ that
% maps function names to a list of variables and the statement that
% defines the function body.  It is used in the rule \textsc{S:Call}.
% }{}
%
% The intuitive meaning of an evaluation \iffull{judgement}{} $(\state,
% S, \cont, \cost) \smallstep{M}^* (\state', S', \cont', \cost')$ is the
% following.  If the statement $S$ is executed in program state
% $\state$, with continuation $\cont$, and with $\cost$ resources
% available then---after a finite number of steps---the evaluation will
% reach the new machine state $(\state', S', \cont',\cost')$ and there
% are $\cost'$ resources available.  If $\cost' \ge 0$ then the
% execution did not run out of resources and the resource consumption up
% to this point is $\cost - \cost'$.  If this difference is negative
% then resource became available during the execution.  If however
% $\cost' < 0$ then the execution ran out of resources and is stuck.
% The cost of the execution is then $\cost \ge 0$.
%
% \iffull{
% \pref{lem:sem1} summarizes the main properties of the resource
% counter of the semantics.  It can be proved by induction on the number
% of steps.
% \begin{lemma}
% \label{lem:sem1}
% \itemskip
%   \begin{enumerate}
%   \item If $(\state, S, \cont, \cost) \smallstep{M}^k (\state', S',
%     \cont', \cost')$ and $n\ge 0$ then $(\state, S, \cont, \cost {+} n)
%     \smallstep{M}^k (\state', S', \cont', \cost' {+} n)$.
% \itemskipIn
%   \item If $(\state, S, \cont, \cost) \smallstep{M}^k (\state', S',
%     \cont', \cost')$ and $(\state, S, \cont, d) \smallstep{M}^k (\state', S',
%     \cont', d')$ then $\cost - \cost' = d - d'$.
%   \end{enumerate}
% \itemskip
% \end{lemma}
% %
% If we assume that we have a \emph{positive resource metric}, that is,
% a metric that assigns a positive cost to each evaluation step then we
% can link bounds on the resource consumption under the metric to
% termination.  This is formalized by \pref{lem:sem_term}.
% %
% \begin{lemma}
%  \label{lem:sem_term}
%  Let $M$ be a positive metric and assume that there exists $b \in
%  \Qplusz$ such that $(\state, S, \cont, \cost) \smallstep{M}^*
%  (\state', S', \cont', \cost') \implies \cost {-} \cost' {\leq} b$.
%  Then there exists $B \in \N$ such that $(\state, S, \cont, d)
%  \smallstep{M}^n (\state'', S', \cont', d')$ for some $n$, $d$, $d'$
%  and $\state''$ implies $n \le B$.
% \end{lemma}
% }

\subsection{Quantitative Hoare Logic}

To prove the soundness of \toolname we found it useful to go through
an intermediate step using a Quantitative Hoare Logic.  This logic
is at the same time a convenient semantic tool and a clean way to
interface manual proofs with our automation, as explained in XXX.
The logic is an extension of the one presented in \cite{veristack14},
it adds support for arbitrary resources and simplifies the
handling of auxiliary state.

Following the previous work, we define quantitative Hoare triples
as $\Delta;B;R\vdash\htriple Q S {Q'}$ where $B$, $R$, $Q$, and $Q'$ are
maps from program states to $\Qplusz\cup\{\infty\}$ in order to represent
an amount of resources available.  The assertions $B$ and $R$ are
postconditions for the case in which the block $S$ exits by a \code{break}
or \code{return} statement.  Additionally, $R$ depends on an extra value
that is returned by the function.  The meaning of the triple $\htriple Q S
{Q'}$ is as follows: If $S$ is executed with starting state $\state$,
the empty continuation \Kstop{} and at least $Q(\state)$ resource units
available then the evaluation does not run out of resources and there
are at least $Q(\state')$ resources left if the evaluation terminates
in state $\state'$.  The logic rules are similar to the ones in previous
work and are simply extended to account for all the costs found in
our cost-aware semantics.

\subsection{The Soundness Theorem}

We translate judgements of the automatic amortized analysis as defined in
\pref{fig:auto} into quantitative Hoare triples.  We define a translation
function $\tr$, such that if a judgement $J$ in the automatic analysis
is derivable, $\tr(J)$ is derivable in the quantitative logic.  By using
$\tr$ to translate derivations of the automatic analysis to derivations
in the quantitative logic we can automatically obtain a certified resource
bound for the analyzed program.

The translation of an assertion $(\Gamma, Q)$ in the automatic
analysis is defined by
$$
  \tr(\Gamma, Q) := \lambda\state .\, (\Gamma(\state)) + \Phi_Q(\state),
$$
where we write $\Phi_Q$ for the unique linear potential function
defined by the quantitative annotation $Q$.  We also need to
translate the assumptions in function contexts of the automatic
analysis.  We define $\tr(\Gamma_f, Q_f, \Gamma'_f, Q'_f) := \forall z \, \vec v \, v . $
%
$$
(
  \lambda \vec v \, \state .\, \tr(\Gamma_f, Q_f)(\state[\Vargs {\mapsto} \vec v]),
  \lambda v \, \state .\, \tr(\Gamma'_f, Q'_f)(\state[\Vret {\mapsto} v])
  )
$$
%
These definitions let us translate the judgement $J = B; R; P \vdash
S \dashv P'$ in the context $\Delta$ by $\tr(J) :=$
$$
  (\lambda f .\, \tr(\Delta(f))); \tr(B); \tr(R) \vdash
    \htriple{\tr(P)}{S}{\tr(P')}.
$$
The soundness of the automatic analysis can now be stated
formally with the following theorem.
%
\begin{theorem}[Soundness of the automatic analysis]
  If $J$ is a judgement derived with the rules of
  \pref{fig:auto}, then $\tr(J)$ is a quantitative
  Hoare triple derivable with the rules of
  \pref{fig:logic}.
\end{theorem}
%
\noindent
The proof of this theorem is constructive and basically maps each rule
of the automatic analysis directly to its counterpart in the
quantitative logic.  The most tricky parts are the translations of the
rules for increments and decrements and the rule {\sc Q:Weak} for
weakening because we have to show that the preconditions of the rules
{\sc L:Weak} or {\sc L:Update}, respectively, are met. %  \iffull{ For instance
% we prove the following lemma to show the soundness of the translation
% of {\sc Q:Weak}.
%
% \begin{lemma}[Relax]
%   If $Q' \succeq_\Gamma Q$ and $\state$ is a program state such that
%   $\state \models \Gamma$ then we have $\Phi_{Q'}(\state) \ge \Phi_Q(\state)$.
% \end{lemma}
% \begin{proof}
% We observe the following inequations.
% \begin{align*}
% \Phi_{Q'}(\state) &= \sum_i q'_i f_i(\state) \\
% &\ge q_0 + \sum_{i\in\mathcal U} u_i p_i - \sum_{i\in\mathcal L} l_i p_i
%   + \sum_{i\in\mathcal U} (q_i - p_i) f_i(\state)  \\ &
%   \;\;\;\;\;\;\;\; + \sum_{i\in\mathcal L} (q_i + p_i) f_i(\state)
%   + \sum_{i\not\in \mathcal U \cup \mathcal L} q_i \\
%  &\ge \Phi_Q(\state)
%   + \sum_{i\in\mathcal U} (u_i - f_i(\state)) p_i
%   + \sum_{i\in\mathcal L} (f_i(\state) - l_i) p_i \\
%  &\ge \Phi_Q(\state).
% \end{align*}
% Since $\state \models \Gamma$, using the transitivity of $\models$ and
% the definition of $\mathcal U$ and $\mathcal L$ we get
% $\forall i\in\mathcal L.\, l_i \le f_i(\state)$ and
% $\forall i\in\mathcal U.\, f_i(\state)\le u_i$.
% These two inequalities justify the last step.
% \end{proof}
% %
% \noindent
% We prove similar lemmas for the rules that deal with assignment.  With
% the computational content of the proof, we have an effective algorithm
% to construct certificates for upper bounds derived automatically.
% }{}

@@
